Two applications of logic programming to Coq

### 2021-02-20 papers/types2020/revisions/app.tex
starting a revision dir for types20. Initial work on sec 3

In this appendix we list some definitions and pieces of code that have been mentioned in the main paper.

@@

\begin{figure}
% \newcommand{\XXi}{\Pscr}
% \newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
% \[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
% \infer{\XXi\vdash G_1\wedge G_2}
% {\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
% \]
% \vskip -18pt
% \[
% \infer{\XXi\vdash A}
%       {\bc D A \quad D \in \Pscr}\qquad
%   \infer{\bc A A}{}{}\qquad
%   \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
%   \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
% \]
% In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
% formulas.
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{code/dep-kernel.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{code/dep-kernel-v2.mod}
\caption{Encoding in coq-elpi of the vanilla meta-interpreter}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}

>>>>

In this appendix we list some definitions and pieces of code that we have  mentioned in the main paper.

@@

\begin{figure}
% \newcommand{\XXi}{\Pscr}
% \newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
% \[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
% \infer{\XXi\vdash G_1\wedge G_2}
% {\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
% \]
% \vskip -18pt
% \[
% \infer{\XXi\vdash A}
%       {\bc D A \quad D \in \Pscr}\qquad
%   \infer{\bc A A}{}{}\qquad
%   \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
%   \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
% \]
% In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
% formulas.
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{../code/dep-kernel.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={interp-end}]{../code/dep-kernel-v2.mod}
\caption{Encoding in coq-elpi of the vanilla meta-interpreter}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}

### 2021-02-20 papers/types2020/revisions/conc.tex
starting a revision dir for types20. Initial work on sec 3

Another obvious limitation of our PBT tactic is that it does not 
 handle any computation (conversion) within an inductive relation. We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
% Mention \emph{Harpoon}, Beluga's new tactic language?

>>>>

Another current limitation of our PBT tactic is that it does not
handle conversion within an inductive relation. One way to address
this is to have the \emph{init} rule use Coq's unification rather than
\lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
%We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
% Mention \emph{Harpoon}, Beluga's new tactic language?

### 2021-02-20 papers/types2020/revisions/cultures.tex
starting a revision dir for types20. Initial work on sec 3

Gentzen's discovered that the key to treating classical and
intuitionistic logics in the same proof system was the weakening and
contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This discovery lead him away from natural
deduction to multiple-conclusion sequent calculus, a very important
kind of proof system that has not been given a satisfactory
representation in type theory.  This same innovation of Gentzen also
opened the way to another key proof-theoretic discovery, that of
linear logic \cite{girard87tcs}.  Similarly, sequent calculus provides
an elegant presentation of linear logic while most treatments of
linear logic in type theory have difficulty treating linear negation
and the multiplicative disjunction $\lpar$ and its unit $\bot$.

>>>>

Gentzen's discovered that the key to treating classical and
intuitionistic logics in the same proof system was the weakening and
contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This discovery lead him away from natural
deduction to multiple-conclusion sequent calculus, a very important
kind of proof system that has not been given a satisfactory
representation in type theory.  This same innovation of Gentzen also
opened the way to another key proof-theoretic discovery, that of
linear logic \cite{girard87tcs}.  Similarly, sequent calculus provides
an elegant presentation of linear logic while most treatments of
linear logic in type theory have difficulty treating linear negation
and the multiplicative disjunction  and its unit.

### 2021-02-20 papers/types2020/revisions/elab.tex
starting a revision dir for types20. Initial work on sec 3

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
\caption{The FPC definition of De Bruijn notation as proof evidence.}
\label{fig:debruijn}
\end{figure}

>>>>

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{../code/deb-fpc.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{../code/deb-fpc.mod}
\caption{The FPC definition of De Bruijn notation as proof evidence.}
\label{fig:debruijn}
\end{figure}

### 2021-02-20 papers/types2020/revisions/pbt.tex
starting a revision dir for types20. Initial work on sec 3

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is
whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to height
at most $5$  from the library \lsti{Inductive} definitions of
\lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
since by mode information we know that it will be computed.
%
Since we did slip in an error (where?), our tactic reports a
counter-example, namely \verb|Proof Term: [0, [0; 1; 0]]|, and
we therefore abort the proof.

In order to generate the PBT query, some
pre-processing is needed. In particular, we turn the data
eigenvariables into \lP logic variables when they appear inside a
specification, and  generate queries %to \lsti|check|
for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to realize this pre-processing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses~\cite{miller91jlc}.

>>>>

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x :: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is
whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall (x : nat) xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to height
at most $5$  from the library \lsti{Inductive} definitions of
\lsti{nat} and \lsti{list}. We do not generate values for \lsti{rs},
since by mode information we know that it will be computed.
%
Since we did slip in an error, our tactic reports a counter-example,
namely \verb|Proof Term: [0, [0; 1; 0]]|, which unpacks to \lsti{x = 0}
and \lsti{xs = [0; 1; 0]}. As the latter is definitely not an
ordered list, this points to a quite evident bug in the definition of
\lsti{ordered}. We leave it to the reader and we abort the proof.

In order to generate the PBT query, some
pre-processing is needed. In particular, we turn 
variables inhabiting datatypes into \lP logic variables when they appear inside a
specification, and  generate queries %to \lsti|check|
for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to realize this pre-processing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses~\cite{miller91jlc}.

### 2021-02-20 papers/types2020/revisions/sec3.tex
starting a revision dir for types20. Initial work on sec 3

A \emph{Horn clause} is a formula of the form $\forall \bar
x_1. A_1\supset \forall \bar x_2. A_2\supset \forall \bar
x_n. A_n\supset A_0$ where $\forall\bar x_i$ denote a list of
universal quantifiers ($i\in\{1,\ldots,n\}$) and $A_0,\ldots,A_n$ are
atomic formulas.  It is well known that the following set of sequent
calculus proof rules are complete for both classical and
intuitionistic logic when one is attempting to prove that a given
atomic formula $A$ is provable from a give set $\Pscr$ of Horn
clauses. 
\[
  \infer[decide]{\Pscr\vdash A}
                {D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\qquad \Pscr\Downarrow D\vdash A}
\]
Here, we use two different styles of sequents.  The sequent
$\Pscr\vdash A$ is the usual sequent which we generally use as the end
sequent (the conclusion) of a proof.  The sequent $\Pscr\Downarrow
D\vdash A$ is a \emph{focused} sequent in which the formula $D$ is the
\emph{focus} of the sequent.  The two left introduction rules and the
initial rule can only be applied to the focused formula.  This latter
point is in contrast to Gentzen's sequent calculus where these rules
can involve \emph{any} formula on the left of the $\vdash$.  The fact
that this proof system is complete for both classical and
intuitionistic logic (when restricted to the Horn clause fragment)
follows from rather simple considerations of Horn clauses
\cite{nadathur90jacm} and from the completeness of LJT proofs
\cite{Herbelin94}.  The use of the term \emph{focus} comes from
Andreoli's proof system for linear logic \cite{andreoli92jlc}.

\begin{figure}[t]
\[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A : T)\in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  A : s}{{E[] \vdash_{CIC} A : s}}% The order should be s : A, not A : s ?
  %% we are checking nat : Set, for example, so A : s --AM
  \qquad
  \infer{\bc{\XXi}{}{[]} A A}{\initLExpert{\XXi}}
  \qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\qquad
         \bc{\XXi_2}{}l {D[t/x]} A\qquad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}

@@

The proof system is parameterized by Coq's global environment $E$,
which here we take as a set of inductive definitions; following Coq's
reference manual, we denote such a definition with
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the inductive type and $\Gamma_C$ the names
and types of its constructors. A rule delegates to Coq's
type checking the enforcement of the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with
types that correspond to Horn clauses, and atomic types are
inductively defined. In fact, we do not have a $\forall$ rule on the
right, although the proof theory would gladly allow it. This means at
the term level that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors of an \lsti{Inductive} definition from the global
environment, one that matches the atom we wish to backchain on, and then
call the latter judgment. The rules for backchaining are standard,
corresponding to the \emph{init} and the (conflation of the) left introduction rules for
$\forall$ and $\supset$.

@@

\begin{figure}[t]
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{code/fpcs.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{code/fpcs.mod}
\caption{Sample FPCs}
\label{fig:resources}
\end{figure}

>>>>

A \emph{Horn clause} is a formula of the form $\forall \bar
x_1. A_1\supset \forall \bar x_2. A_2\supset \forall \bar
x_n. A_n\supset A_0$ where $\forall\bar x_i$ denote a list of
universal quantifiers ($i\in\{1,\ldots,n\}$) and $A_0,\ldots,A_n$ are
atomic formulas.  It is well known that the following set of sequent
calculus proof rules are complete for both classical and
intuitionistic logic when one is attempting to prove that a given
atomic formula $A$ is provable from a  set $\Pscr$ of Horn
clauses. 
\[
  \infer[decide]{\Pscr\vdash A}
                {D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\qquad \Pscr\Downarrow D\vdash A}
\]
Here, we use two different styles of sequents.  The sequent
$\Pscr\vdash A$ is the usual sequent which we generally use as the end
sequent (the conclusion) of a proof.  The sequent $\Pscr\Downarrow
D\vdash A$ is a \emph{focused} sequent in which the formula $D$ is the
\emph{focus} of the sequent.  The two left introduction rules and the
initial rule can only be applied to the focused formula.  This latter
point is in contrast to Gentzen's sequent calculus where these rules
can involve \emph{any} formula on the left of the $\vdash$.  The fact
that this proof system is complete for both classical and
intuitionistic logic (when restricted to the Horn clause fragment)
follows from rather simple considerations of Horn clauses
\cite{nadathur90jacm} and from the completeness of LJT proofs
\cite{Herbelin94}.  The use of the term \emph{focus} comes from
Andreoli's proof system for linear logic \cite{andreoli92jlc}.

\begin{figure}[t]
\[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A) : T \in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  A : s}{{E[] \vdash_{CIC} A : s}}% The order should be s : A, not A : s ?
  %% we are checking nat : Set, for example, so A : s --AM
  \qquad
  \infer{\bc{\XXi}{}{[]} A A}{\initLExpert{\XXi}}
  \qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\qquad
         \bc{\XXi_2}{}l {D[t/x]} A\qquad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{../code/dep-kernel.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{../code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}

@@

The proof system is parameterized by Coq's global environment $E$,
which here we take as a set of inductive definitions; following Coq's
reference manual, we denote such a definition with
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the (possibly mutually)inductive type and
$\Gamma_C$ the names and types of its constructors; finally $p$
denotes the number of parameters and playes no role. We delegate to
Coq's type checking the enforcement of the well-sortedness of
inductive types. The \emph{local context} is empty, since we are only
dealing with types that correspond to Horn clauses, and atomic types
are inductively defined. In fact, we do not have a $\forall$ rule on
the right, although the proof theory would gladly allow it. This means
at the term level that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors of an \lsti{Inductive} definition from the global
environment, one that matches the head symbol of the atom we aim to backchain on, and then
call the latter judgment that will find a correct instantiation, if any. The rules for backchaining are standard,
corresponding to the \emph{init} and the (conflation of the) left introduction rules for
$\forall$ and $\supset$.

@@

\begin{figure}[t]
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{../code/fpcs.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={resources-end}]{../code/fpcs.mod}
\caption{Sample FPCs}
\label{fig:resources}
\end{figure}
